Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-14 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Spell check, minor corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2874 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 829528fe7fe5d19d183731dd82fe3532ff29c345

To compare the predictive performances of all proposed models a circular
analogue of the continuous ranked probability score (CRPS) as introduced by
\cite{Grimit+Gneiting+Berrocal:2006} is computed. Just as the linear version of
the CRPS \citep[for more details see][]{Hersbach:2000} it is a proper scoring
rule \citep{Gneiting+Raftery:2007} and measures the difference between an
observation and the corresponding predicted distribution function in order to
assess the probabilistic goodness of fit for the estimated model. Hence, the lower
the CRPS value the better the performance.  Contrary to the linear version, the
circular CRPS reduces not to the absolute error but to the angular distance
when the forecast is deterministic.

In addition to the raw CRPS, to assess differences in the
improvement of the various statistical models over the climatological model
used as a reference method, CRPS-based skill scores are computed according to
\begin{equation} 
  \text{CRPSS}_{\text{method}} = 1 -
\frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{climatology}}}.
\end{equation}

All scores presented in the next section are out-of-sample: For the persistency
model only time points prior to the time of interest are used, for all other
models a five-fold cross-validation is employed using up to four calendar years
for model training and the remaining single calendar year for validation. Due to the large sample
size of 24-hourly values per day over five years, some kind of temporal
aggregation is needed to ensure a correct visual comparison of the individual
methods. As for the employed models the variablilty of the predictive performance over the five years
is lower than over a single day or over a single year, CRPS and CRPS skill scores are
aggregrated over the respective five validation years; this yields 24 hourly
scores per 12 months averaged over the five validation years.

\subsection{Results} 
This section provides a detailed analysis on the predictive performance of the
different proposed statistal models applied to probabilistic wind direction
forecasting. To ensure a comprehensive comparison of the models, wind direction
forecasts are evaluated for two different lead times at two airports with
different climatological site characteristics.
Figure~\ref{fig:boxplot_crpsraw} shows the CRPS values of the employed models
at forecast steps 1\,h and 3\,h for the airports Innsbruck (panels a,\,c) and
Vienna (panels b,\,d). The scores are aggregated over the $5$~validation years,
yielding yearly mean values for every hour per calendar month, with a lower
score indicating better performance.  For both stations at both forecast steps,
the circular regression forest overall provides the best predictive
performance, followed by the circular regression tree and, except for the
$3$\,h forecast at Innsbruck, the persistency model.  In comparison to the
circular tree and forest, for both stations and forecast steps, the
climatological model and the linear model show clearly higher CRPS scores and
hence a lower predictive performance.

The different site characteristics between the airports Innsbruck
(Figure~\ref{fig:boxplot_crpsraw}a,\,c) and Vienna
(Figure~\ref{fig:boxplot_crpsraw}b,\,d) seem to have an effect on the absolute
level of the respective model performances and on their respective performance
variation. At Innsbruck, due to the surrounding mountains a limited number of
possible wind directions exists, namely the three wind regimes discussed for
Figure~\ref{fig:tree_ibk} in Section~\ref{sec:introduction}. Therefore, for
Innsbruck the wind direction remains rather constant in one of these possible
states, but once a change takes place it is mostly a major wind direction
shift.  Hence, due to the few wind regimes the rather unflexible climatological
and linear models perform relatively similar to the other models
(Figure~\ref{fig:boxplot_crpsraw}a,\,c). In addition, at Innsbruck the potential
high prediction errors in case of a paradigma change seem to lead to an higher
variation in the predictive performance for all models in comparison to Vienna;
this variation is especially high for the persistency model due to its strong
vulnerability to abrupt wind shifts. On the contrary, at Vienna the geographic
surroundings do not constrain any wind direction which leads to smaller and
less abrupt changes in the wind direction as well as less pronounced wind
regimes. This seems to weaken the predictive performance of the climatological
and linear models, and to reduce the performance variability for all models
(Figure~\ref{fig:boxplot_crpsraw}b,\,d).

The different forecast steps seem to have only a minor effect on the predictive
performance of the climatological model and the linear model for both stations.
However, for the persistency model, at both stations, higher scores for the
$3$\,h forecast (panels c,\,d) reveal a lower performance for longer lead times. This is
clearly due to at least 3-hourly instead of 1-hourly lagged response values
employed as covariates in the persistency model. Apparently, the circular
regression tree and forest can partially compensate the lower skill of
the lagged response values by other covariates, hence their predictive
perfomance only slightly decreases for the 3-hourly in comparison to the
1-hourly wind direction forecast. This compensation is especially evident for
Innsbruck, where the performance difference between the persistency model and
the tree-based methods significantly increases for the 3-hourly forecast, with
a significantly better performance for the circular tree and forest.

@@

In addition to the raw CRPS, Figure~\ref{fig:boxplot_crpsskill} provides CRPS
skill scores with the climatological model as a reference. Skill scores are in
percent, and positive values indicate an improvement in the predictive
performance over the reference. For all setups, the circular regression forest
has the highest skill scores with a mean performance gain of 13--25\% and 58--71\% for
Innsbruck and Vienna, respectively. As discussed for
Figure~\ref{fig:boxplot_crpsraw}, this improvement over the climatological
model is lower for Innsbruck due to the few number of predominat wind regimes
and hence a relatively good performance of the climatological model.
Additionally, Figure~\ref{fig:boxplot_crpsskill} shows that even if the
persistency model's performance is lower than the reference (panel c) the
tree-based models are significantly superior to the climatolical model used as
reference.

>>>>

To compare the predictive performances of all proposed models a circular
analogue of the continuous ranked probability score (CRPS) as introduced by
\cite{Grimit+Gneiting+Berrocal:2006} is computed. Just as the linear version of
the CRPS \citep[for more details see][]{Hersbach:2000} it is a proper scoring
rule \citep{Gneiting+Raftery:2007} and measures the difference between an
observation and the corresponding predicted distribution function in order to
assess the probabilistic goodness of fit for the estimated model. Hence, the
lower the CRPS value the better the predictive performance.  Contrary to the
linear version, the circular CRPS reduces not to the absolute error but to the
angular distance when the forecast is deterministic.

In addition to the raw CRPS, to assess differences in the
improvement of the various statistical models over the climatological model
used as a reference, CRPS-based skill scores are computed according to
\begin{equation} 
  \text{CRPSS}_{\text{method}} = 1 -
\frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{climatology}}}.
\end{equation}

All scores presented in the next section are out-of-sample: For the persistency
model only time points prior to the time of interest are used, for all other
models a five-fold cross-validation is employed using up to four calendar years
for model training and the remaining single calendar year for validation. Due
to the large sample size of 24-hourly values per day over five years, some kind
of temporal aggregation is needed to ensure a correct visual comparison of the
individual methods. As for the employed models the variability of the
predictive performance over the five years is lower than over a single day or
over a single year, CRPS and CRPS skill scores are aggregated over the
respective five validation years; this yields 24 hourly scores per 12 months
averaged over the five validation years.

\subsection{Results} 
This section provides a detailed analysis on the predictive performance of the
different proposed statistical models applied to probabilistic wind direction
forecasting. To ensure a comprehensive comparison of the models, wind direction
forecasts are evaluated for two different lead times at two airports with
different climatological site characteristics.
Figure~\ref{fig:boxplot_crpsraw} shows the CRPS values of the employed models
at forecast steps 1\,h and 3\,h for the airports Innsbruck (panels a,\,c) and
Vienna (panels b,\,d). The scores are aggregated over the $5$~validation years,
yielding yearly mean values for every hour per calendar month, with a lower
score indicating better performance.  For both stations at both forecast steps,
the circular regression forest overall provides the best predictive
performance, followed by the circular regression tree and, except for the
$3$\,h forecast at Innsbruck, the persistency model.  In comparison to the
circular tree and forest, for both stations and forecast steps, the
climatological model and the linear model show clearly higher CRPS scores and
hence a lower predictive performance.

The different site characteristics between the airports Innsbruck
(Figure~\ref{fig:boxplot_crpsraw}a,\,c) and Vienna
(Figure~\ref{fig:boxplot_crpsraw}b,\,d) seem to have an effect on the absolute
level of the respective model performances and on their respective performance
variation. At Innsbruck, due to the surrounding mountains a limited number of
possible wind directions exists, namely the three wind regimes discussed for
Figure~\ref{fig:tree_ibk} in Section~\ref{sec:introduction}. Therefore, for
Innsbruck the wind direction remains rather constant in one of these possible
states, but once a change takes place it is mostly a major wind direction
shift.  Hence, due to the few wind regimes the rather inflexible climatological
and linear models score relatively well with similar values as the other models
(Figure~\ref{fig:boxplot_crpsraw}a,\,c). In addition, at Innsbruck the
potential high prediction errors in case of a paradigm shift seem to lead to
an higher variation in the predictive performance for all models in comparison
to Vienna; this variation is especially high for the persistency model due to
its strong vulnerability to abrupt wind shifts. On the contrary, at Vienna the
geographic surroundings do not constrain any wind direction which leads to
smaller and less abrupt changes in the wind direction as well as less
pronounced wind regimes. This seems to weaken the predictive performance of the
climatological and linear models, and to reduce the performance variability for
all models (Figure~\ref{fig:boxplot_crpsraw}b,\,d).

The different forecast steps seem to have only a minor effect on the predictive
performance of the climatological model and the linear model at both stations.
However, for the persistency model, at both stations, higher scores for the
$3$\,h forecast (panels c,\,d) reveal a lower performance for longer lead
times; this is probably due to minimum 3-hourly instead of 1-hourly lagged
response values employed as covariates in the persistency model. The circular
regression tree and forest seem to partially compensate for the lower skill of
the lagged response values by other covariates, hence their predictive
performance only slightly decreases for the longer lead time.  This compensation
is especially evident for Innsbruck, where the performance difference between
the persistency model and the tree-based methods significantly increases from
the 1-hourly to the 3-hourly forecast.

@@

In addition to the raw CRPS, Figure~\ref{fig:boxplot_crpsskill} provides CRPS
skill scores with the climatological model as a reference. Skill scores are in
percent, and positive values indicate an improvement in the predictive
performance over the reference. For all setups, the circular regression forest
has the highest skill scores with a mean performance gain of 13--25\% and 58--71\% for
Innsbruck and Vienna, respectively. As discussed for
Figure~\ref{fig:boxplot_crpsraw}, this improvement over the climatological
model is lower for Innsbruck due to the few number of predominant wind regimes
and hence a relatively good performance of the climatological model.
Additionally, Figure~\ref{fig:boxplot_crpsskill} shows that even if the
persistency model's performance is lower than the reference (panel c) the
tree-based models are significantly superior to the climatological model used as
reference.
